Ethnicity sensitive author disambiguation using semi-supervised learning

### 2016-04-29 paper/author_disambiguation.tex
kesw: first official version
Signed-off-by: Mateusz Susik <mateuszsusik@gmail.com>

\usepackage{xspace}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{cite}
\usepackage[hidelinks]{hyperref}
\newcommand*{\eg}{e.g.\@\xspace}
\newcommand*{\ie}{i.e.\@\xspace}
\usepackage{color}
\usepackage{microtype}
\setlength{\intextsep}{5pt plus 0pt minus 0pt}
\newcommand{\msnote}[1]{\textcolor{blue}{[MS: #1]}}

\def\baselinestretch{0.98}

\newcommand{\longpage}{\enlargethispage{\baselineskip}}

@@

The remainder of this report is structured as follows. In Section~\ref{related-works},
we briefly review machine learning solutions for author disambiguation. The components of our method are then defined in Section~\ref{methods} and its implementation described in Section~\ref{implementation}. Experiments
are carried out in Section~\ref{experiments}, where we explore and validate
features for the supervised learning of a linkage function and compare
strategies for the semi-supervised clustering of publications.
Finally, conclusions and future works are discussed in Section~\ref{conclusions}.

@@

Due to the lack of large and publicly available datasets of curated
clusters of publications, studies on author disambiguation are usually
constrained to validating their results on manually built datasets of limited
size and scope (from a few hundred to a few thousand papers, with sparse
coverage of ambiguous cases), making the true performance of these methods
often difficult to assess with high confidence.
Additionally, despite devoted efforts to construct them, these datasets are rarely public,
making it even more difficult to compare methods using a common benchmark.

@@

To form flat clusters from the hierarchy, one must decide on a maximum
distance threshold above which clusters are considered to correspond to
distinct authors.
Let us denote by ${\cal S}^\prime = \{ s | s \in c^\prime, c^\prime
\in {\cal C}^\prime \}$ the set of all signatures for which partial clusters are
known.
Let us also denote by $\smash{\widehat{\cal C}}$ the predicted clusters for all signatures in ${\cal S}$, and by
$\smash{\widehat{\cal C}^\prime} = \{ \widehat{c} \cap {\cal S}^\prime | \widehat{c} \in \widehat{\cal C} \}$
the predicted clusters restricted to signatures for which partial clusters are known.
From these, we evaluate the following semi-supervised cut-off strategies, as illustrated in Figure~\ref{fig:cuts}:
\begin{itemize}

@@

\subsubsection{Baseline.} The simplest baseline against which we compare our results
consists in grouping all signatures sharing the same (normalized) surname(s) and
the same (normalized) first given name initial. It provides a simple and fast
solution yielding decent results, as reported at the top of
Table~\ref{table:results}

\begin{small}
\begin{table*}
\caption{Average precision, recall and f-measure scores on test folds. Underlined components correspond to the state-of-the-art choices.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c | c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}}\\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$ \\
  \hline
  \hline
Baseline & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
\hline
\underline{Blocking = SFI} & 0.9901 & 0.9760 & 0.9830  & 0.9948 & 0.9738 & 0.9842\\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841  & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850}  & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
No name normalization & 0.9887 & 0.9697 & 0.9791 & 0.9931 & 0.9658 & 0.9793 \\
\underline{Name normalization} & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842}  \\
\hline
\underline{Classifier = GBRT} & 0.9901 & 0.9760 & 0.9830   & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846}  & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711  & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786  & 0.9850 & 0.9707 & 0.9778 \\
\underline{Training pairs = Blocked, balanced} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\underline{Clustering = Average linkage} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671  & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785  & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut (baseline) & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814  & 0.9940 & 0.9727 & 0.9832 \\
\underline{Block cut} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Combined best settings & 0.9897 & 0.9827 & \textbf{0.9862}  & 0.9954 & 0.9805 & \textbf{0.9879} \\
Best settings without ethnicity features & 0.9862 & 0.9819 & 0.9841 & 0.9937 & 0.9815 & 0.9876 \\
  \hline
\end{tabular}
\end{table*}
\end{small}

@@

\begin{itemize}
	\item Blocking: same surname and the same first given name initial strategy (SFI) and with name normalization;
	\item Linkage function: all 22 features defined in Table~\ref{table:features},
	gradient boosted regression trees as supervised learning algorithm
	and a training set of pairs built from $({\cal S}_\text{train}^\prime, {\cal C}_\text{train}^\prime)$, by balancing easy and difficult cases.
	\item Clustering: agglomerative clustering using average linkage and
	block cuts found to maximize $F_\text{B3}({\cal C}_\text{train}^\prime, \widehat{\cal C}_\text{train}^\prime, {\cal S}_\text{train}^\prime)$.
\end{itemize}

@@

Finally, as discussed previously, the step of
normalizing author names (stripping accents, removing affixes), as performed in the
state-of-the-art, is shown to be important. Results from Table~\ref{table:results} clearly suggest that not
normalizing significantly reduces performance (yielding an F-measure of $0.9830$ when normalizing,
but decreasing to $0.9791$ when raw author name strings are used instead).

@@

Moreover, we observed that the ethnicity features are an improvement. When these features were not
included in the features set, using the best combined settings, the algorithm yields worse result
($0.9841$).

@@

Clustering together all signatures from the same block (\ie, baseline) is the least effective
strategy ($0.9409$), but yields anyhow surprisingly decent accuracy, given the
fact it requires no linkage function and no agglomerative clustering -- only the blocking
function is needed to group signatures. In particular, this result reveals
that author names are not ambiguous in most cases\footnote{This holds for the
data we extracted, but may in the future, with the rise of non-Western
researchers, be an underestimate of the ambiguous cases.} and that only a small
fraction of them requires advanced disambiguation procedures. On the other
hand, both global and block cut thresholding strategies give better results,
with a slight advantage for the block cuts ($0.9814$ versus $0.9830$), as expected.
In case ${\cal S}^\prime_b$ is empty (\ie partial clusters are not known for any of the signatures from the
block), this therefore suggests that either using
a cut-off threshold learned globally from the known data would in general give results only
marginally worse than if the claimed signatures had been
known. Trying out 3 clustering methods, the results show that \textit {average linkage} was the best one.

@@

\subsubsection{Combined best settings} When all best settings are combined (\ie, Blocking = NYSIIS, Name normalization, Classifier = Random Forests, Training
pairs = blocked and balanced, Clustering = Average linkage, Block cuts),
performance reaches $0.9862$, \ie, the best of all reported results. In particular,
this combination exhibits both the high recall of phonetic blocking based on the NYSIIS algorithm and the high precision of Random Forests.

@@

Overall, these results all encourage further improvements and research. For
blocking, one of the open challenges is to manage signatures with inconsistent
surnames or inconsistent first given names (cases 4 and 5, as described in
Section~\ref{methods:blocking}) while maintaining blocks to a tractable size.
As phonetic algorithms are not yet perfect, another direction  for further work is the design of better
phonetization functions, tailored for author disambiguation. For the linkage function,
the good results of the proposed features pave the way for further research  in
ethnicity-sensitive author disambiguation. The automatic fitting of the
pipeline to cultures and ethnic groups for which standard author disambiguation
is known to be less efficient (\eg, Chinese authors with many homonyms)  indeed
constitutes a direction of research with great potential benefits for the
concerned scientific communities.

@@

>>>>

\usepackage{xspace}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{cite}
\usepackage[hidelinks]{hyperref}
\newcommand*{\eg}{e.g.\@\xspace}
\newcommand*{\ie}{i.e.\@\xspace}
\usepackage{color}
\usepackage{microtype}

\setlength{\intextsep}{5pt plus 0pt minus 0pt}
\def\baselinestretch{0.99}

\newcommand{\longpage}{\enlargethispage{\baselineskip}}
\newcommand{\verylongpage}{\enlargethispage{2\baselineskip}}

@@

The remainder of this report is structured as follows. In Section~\ref{related-works},
we briefly review machine learning solutions for author disambiguation. The components of our method are then defined in Section~\ref{methods} and its implementation described in Section~\ref{implementation}. Experiments
are carried out in Section~\ref{experiments}, where we compare
approaches to the problem and explore feature choice.
Finally, conclusions and future works are discussed in Section~\ref{conclusions}.

@@

%Due to the lack of large and publicly available datasets of curated
%clusters of publications, studies on author disambiguation are usually
%constrained to validating their results on manually built datasets of limited
%size and scope (from a few hundred to a few thousand papers, with sparse
%coverage of ambiguous cases), making the true performance of these methods
%often difficult to assess with high confidence.
%Additionally, despite devoted efforts to construct them, these datasets are rarely public,
%making it even more difficult to compare methods using a common benchmark.

@@

To form flat clusters from the hierarchy, one must decide on a maximum
distance threshold above which clusters are considered to correspond to
distinct authors.
Let us denote by ${\cal S}^\prime = \{ s | s \in c^\prime, c^\prime
\in {\cal C}^\prime \}$ the set of all signatures for which partial clusters are
known.
Let us also denote by $\smash{\widehat{\cal C}}$ the predicted clusters for all signatures in ${\cal S}$, and by
$\smash{\widehat{\cal C}^\prime} = \{ \widehat{c} \cap {\cal S}^\prime | \widehat{c} \in \widehat{\cal C} \}$
the predicted clusters restricted to signatures for which partial clusters are known.
From these, we evaluate the following semi-supervised cut-off strategies, as illustrated in Figure~\ref{fig:cuts}:

\begin{itemize}

@@

\subsubsection{Baseline.} The simplest baseline against which we compare our results
consists in grouping all signatures sharing the same (normalized) surname(s) and
the same (normalized) first given name initial. It provides a simple and fast
solution yielding decent results, as reported at the top of
Table~\ref{table:results}.

\begin{small}
\begin{table*}
\caption{Average precision, recall and f-measure scores on test folds. - components correspond to the state-of-the-art choices.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c | c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}}\\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$ \\
  \hline
  \hline
Baseline & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
\hline
\underline{Blocking = SFI} & 0.9901 & 0.9760 & 0.9830  & 0.9948 & 0.9738 & 0.9842\\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841  & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850}  & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
\underline{Classifier = GBRT} & 0.9901 & 0.9760 & 0.9830   & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846}  & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711  & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786  & 0.9850 & 0.9707 & 0.9778 \\
\underline{Training pairs = Blocked, balanced} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
\underline{Clustering = Average linkage} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671  & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785  & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut (baseline) & 0.9024 & 0.9828 & 0.9409  & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814  & 0.9940 & 0.9727 & 0.9832 \\
\underline{Block cut} & 0.9901 & 0.9760 & \textbf{0.9830}  & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Combined best settings & 0.9888 & 0.9848 & \textbf{0.9868}  & 0.9951 & 0.9831 & \textbf{0.9890} \\
Best settings without ethnicity features & 0.9862 & 0.9819 & 0.9841 & 0.9937 & 0.9815 & 0.9876 \\
  \hline
\end{tabular}
\end{table*}
\end{small}

@@

\begin{itemize}
	\item Blocking: same surname and the same first given name initial strategy (SFI);
	\item Linkage function: all 22 features defined in Table~\ref{table:features},
	gradient boosted regression trees as supervised learning algorithm
	and a training set of pairs built from $({\cal S}_\text{train}^\prime, {\cal C}_\text{train}^\prime)$, by balancing easy and difficult cases.
	\item Clustering: agglomerative clustering using average linkage and
	block cuts found to maximize $F_\text{B3}({\cal C}_\text{train}^\prime, \widehat{\cal C}_\text{train}^\prime, {\cal S}_\text{train}^\prime)$.
\end{itemize}

@@

%Finally, as discussed previously, the step of
%normalizing author names (stripping accents, removing affixes), as performed in the
%state-of-the-art, is shown to be important. Results from Table~\ref{table:results} clearly %suggest that not
%normalizing significantly reduces performance (yielding an F-measure of $0.9830$ when %normalizing,
%but decreasing to $0.9791$ when raw author name strings are used instead).

@@

Moreover, we observed that the ethnicity features serve a purpose. When these features were not
included in the features set, using the best combined settings, the algorithm yields worse performance
($0.9841$).

@@

@@

Clustering together all signatures from the same block (\ie, baseline) is the least effective
strategy ($0.9409$), but yields anyhow surprisingly decent accuracy, given the
fact it requires no linkage function and no agglomerative clustering -- only the blocking
function is needed to group signatures. In particular, this result reveals
that author names are not ambiguous in most cases\footnote{This holds for the
data we extracted, but may in the future, with the rise of non-Western
researchers, be an underestimate of the ambiguous cases.} and that only a small
fraction of them requires advanced disambiguation procedures. On the other
hand, both global and block cut thresholding strategies give better results,
with a slight advantage for the block cuts ($0.9814$ versus $0.9830$), as expected.
In case ${\cal S}^\prime_b$ is empty (\ie partial clusters are not known for any of the signatures from the
block), this therefore suggests that either using
a cut-off threshold learned globally from the known data would in general give results only
marginally worse than if the claimed signatures had been
known.

\subsubsection{Combined best settings} When all best settings are combined (\ie, Blocking = NYSIIS, Classifier = Random Forests, Training
pairs = blocked and balanced, Clustering = Average linkage, Block cuts),
performance reaches $0.9862$, \ie, the best of all reported results. In particular,
this combination exhibits both the high recall of phonetic blocking based on the NYSIIS algorithm and the high precision of Random Forests.

@@

Overall, these results all encourage further improvements and research. For
blocking, one of the challenges is to manage signatures with inconsistent
surnames or first given names (cases 4 and 5, as described in
Section~\ref{methods:blocking}) while maintaining blocks to a tractable size.
As phonetic algorithms are not yet perfect, another direction  for further work is the design of better
phonetization functions, tailored for author disambiguation. For the linkage function,
the good results of the proposed features pave the way for further research  in
ethnicity-sensitive author disambiguation. The automatic fitting of the
pipeline to cultures and ethnic groups for which standard author disambiguation
is known to be less efficient (\eg, Chinese authors with many homonyms)  indeed
constitutes a direction of research with great potential benefits for the
concerned scientific communities.

@@

\verylongpage
